---
title: "NBA Scorigami Observatory"
format: html
execute:
  echo: false
  warning: false
  error: false
---

```{r}
library(ggplot2)
library(dplyr)
library(lubridate)
library(scales)
library(grid)

games <- read.csv("Games.csv", stringsAsFactors = FALSE)

games <- games |>
  mutate(
    gameDate = as.Date(gameDateTimeEst),
    year = year(gameDate),
    homeScore = as.integer(homeScore),
    awayScore = as.integer(awayScore)
  ) |>
  filter(!is.na(homeScore), !is.na(awayScore))

score_grid <- games |>
  filter(homeScore >= 16, homeScore <= 192,
         awayScore >= 16, awayScore <= 192)

score_trend <- games |>
  group_by(year) |>
  summarise(unique_scores = n_distinct(paste(homeScore, awayScore, sep = "-")), .groups = "drop") |>
  arrange(year) |>
  mutate(cumulative = cumsum(unique_scores))

total_scores <- n_distinct(paste(score_grid$homeScore, score_grid$awayScore, sep = "-"))
unique_years <- length(unique(score_trend$year))
median_home <- median(score_grid$homeScore)
median_away <- median(score_grid$awayScore)

p_main <- ggplot(score_grid, aes(x = homeScore, y = awayScore)) +
  stat_bin2d(binwidth = c(4, 4), colour = NA) +
  scale_fill_gradientn(
    colours = c("#120A31", "#373179", "#5c52a6", "#c67047", "#fecb78", "#fdf1c7"),
    values = rescale(c(0, 0.1, 0.3, 0.6, 0.85, 1)),
    guide = guide_colorbar(barwidth = 12, barheight = 0.5, title = "Density")
  ) +
  geom_abline(slope = 1, intercept = 0, colour = "white", size = 0.6, alpha = 0.85) +
  scale_x_continuous(breaks = seq(20, 192, by = 20), limits = c(16, 192)) +
  scale_y_continuous(breaks = seq(20, 192, by = 20), limits = c(16, 192)) +
  labs(
    x = "Home PTS",
    y = "Visitor PTS",
    title = "NBA Scorigami",
    subtitle = "A heatmap of every game score combination in Games.csv",
    caption = "Data: Games.csv"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.background = element_rect(fill = "#f5f5f8", colour = NA),
    plot.background = element_rect(fill = "#f5f5f8", colour = NA),
    panel.grid.major = element_line(colour = "#d9d9e0"),
    panel.grid.minor = element_blank(),
    axis.text = element_text(colour = "#3c3f56"),
    axis.title = element_text(colour = "#3c3f56", face = "bold"),
    plot.title = element_text(size = 28, face = "bold", colour = "#1f2648"),
    plot.subtitle = element_text(size = 12, colour = "#4b5172"),
    plot.caption = element_text(size = 8, colour = "#6c6f85"),
    legend.position = "bottom",
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8)
  )
```

::: {.hero}
# NBA Scorigami Observatory

Explore the rare and familiar NBA score combinations from the dataset, then zoom into the patterns behind the numbers.

[View the Score Map](#scores){.cta}
:::

::: {.stats-grid}

::: {.stat-card}
## Unique scorelines
`r formatC(total_scores, format = "d", big.mark = ",")`
:::

::: {.stat-card}
## Seasons covered
`r formatC(unique_years, format = "d")`
:::

::: {.stat-card}
## Median home score
`r median_home`
:::

::: {.stat-card}
## Median away score
`r median_away`
:::

:::

::: {.section-card #scores}
## Score heatmap
Each colored cell represents how often that exact home/away score pair has occurred in the dataset. Brighter cells mean more common scorelines.

```{r}
p_main
```
:::

::: {.section-card #insights}
## Insights
What stands out is the dense diagonal band: many games are close finishes. Low-scoring games are rare, while the mid-range buckets around 100-120 points dominate. This dashboard surfaces where NBA score combinations cluster and where the league barely visits.
:::

::: {.footer}
Created with R and Quarto — play with the data by editing `Games.csv` and rerunning this page.
:::
